Bayesian Inference Using Sequential Monte-Carlo Algorithm for Dynamic System Models

### 2020-08-10 background/background.tex
C1 and C2 read

 [describe the process and how cells and cytokines are involved]

Compared to mammals, zebrafish can regenerate spinal cord after injury, and possibly recover swimming function. They are widely used in regenerative ability researches. Experiments found that although being cut off, the axonal of nerve cells can regenerate and bridge across the lesion site, which critically determined whether zebrafish can regain the swimming function \cite{axonal}. Tsarouchas et al. \cite{ref:Tsarouchas} studied the the complex interactions happen at the lesion site. The inflammation affects the regeneration, as if it is promoted, it accelerates the axonal regeneration; if it is inhibited, it reduces the regeneration. Further experiments showed that two kinds of cytokines, il-1$\beta$ and tnf-$\alpha$, affect the inflammation, which are dynamically controlled by immunes cells, i.e. cytokines works as media to control the regeneration. Here cytokines are small proteins, work as messenger to pass signal across different cells, which are similar to hormones and neurotransmitter.

An interaction map can be drawn, according to the evidences found in \cite{ref:Tsarouchas}. After injury, a massive immune response was observed, with the increasing numbers of several kinds of cells: macrophage, neutrophil and keratinocyte. The dynamical system consists of complex interactions between cells and cytokines, e.g. il-1$\beta$ promotes macrophages' increase. and as a major source, more macrophages leads to more expression of tnf-$\alpha$ mRNA. Consequently, tnf-$\alpha$ and il-1$\beta$ plays determinative roles in the axonal regeneration. Generally, il-1$\beta$ promotes the regeneration at early stages, but later has an inhibition effect, and tnf-$\alpha$ promotes the regeneration all the time. The presence if the cytokines also affects the number of cell, while as sources number of cells also affects cytokines' expression. Interactions between different kinds of cells are also considered in the system.

This is a dynamic system suitable for mathematical modelling, as a set of equations can be written to represent the dynamics as possible. Our interest lies in the modelling of the regeneration period of the dynamic system, and we wish to find that if we can quantitatively study the interactions iniside the system, based on the proposed hypothesis \cite{ref:Tsarouchas}, and our own hypothesis which may include different interactions.

@@

 [how this can be modelled as dynamic systems]

 [general topics of mathematical models: how to build model according to interactions; how to calibrate/ parameterise the model; how to solve the model; dynamics; how to evaluate the model]

Mathematical models can express real-life questions in mathematical forms, and provide qualitative or quantitative inside-look of the question through various mathematical tools and techniques. Mathematical models can be used for prediction, classification, testing our understanding or hypothesis and many other tasks, thus they are widely used to solve practical problems. However, for a model that describe the zebrafish spinal cord regeneration process, we were less interested to the model's predictive functionality; the main focus is on the explore of uncertainty in parameter estimates and comparison with alternative models. In this case, we wish to use modelling techniques to test our inference method, our hypothesis and hopefully an well-fitted models can be found and compared.

According to our understanding of the biological mechanism in the regeneration, hypothesis should firstly proposed, in a way which can represent the real  mechanism as much as possible. Usually a real-life problem, e.g. the spinal cord regeneration, consists of complex dynamics and interactions with many other environmental factors to be considered, An elaborate description of the real cases using a mathematical model sometimes can be impossible; assumptions and hypothesis are proposed to make necessary simplification and abstract for a final reasonably detailed model. Some qualitative models are simple but illustrative for the real-life observed patterns, e.g. Lotka-Volterra (L-V) equation for predator and prey models; they are usually analytically solvable and some properties are used to make predictions (e.g. steady states in ecosystem). If a more precise model is desired, usually more effects or interactions are considered and added as terms to a re-constructed model, e.g. environmental factors, a changing environmental capacity and predation from other species.

Models like L-V equation are mostly analytically solvable when analysing properties of the model, however, analytical solution for many models are not feasible, because of more complex dynamics, or the model itself is not a deterministic model. In the areas that are not analytically feasible, usually computational techniques are applied for approximate solutions. Solvers for different models are used to simulate the system or find roots for some spacial points.

For the regeneration process that we wanted to model, differential equations are our preference. Non-linear ordinary differential equations are used in this dissertation, where the only independent variable is time $t$ denoting the post-lesion time (hours), but further models can also consider other models e.g. stochastic models and partial differential equations. For ordinary differential equations (ODE) system, there are several available computational solvers using different algorithms, implemented in MATLAB\footnote[1] {\url{https://uk.mathworks.com/help/matlab/math/choose-an-ode-solver.html}}, Python\footnote[2]{\url{https://docs.scipy.org/doc/scipy/reference/generated/scipy.integrate.solve_ivp.html}} and many other languages. These solvers help to simulate the trajectories of the dependent variables, given initial values and a set of fixed parameters. The goodness of `fit' then can be measured by compared the simulated data and observed real data.

Often further researches on the modelled problems require the model to be deterministic on its parameters. Parameters in the dynamic system models usually have their practical significance, e.g. describing reaction speed, environmental capacity or the birth rate. There are many approaches to derive these parameters from data or experience. For a specific problem and a widely used models, the parameters' values can be read from previous literatures; however in many others cases, we need to find parameter's values specifically for our model and problem, where numerical values for parameters are more suitable for practical analysis using ODE solvers or other tasks.

To find the best parameters' values given observed data, either optimisation-based approaches (e.g. least-square (LS) fit) or inference approaches can be used. Optimisation-based approaches may face two major drawbacks: over-fitting and local optima \cite{ref:abcsysbio}. Bayesian inference approaches gain popularity in recent years, which can relief the concerns in over-fitting and provide a measurement for the uncertainty in the inferred values. Local optima can still be a major concern, and many efforts have be proposed for a more through explore of the parameter space.

@@

 [how to parameterise the model]

 [how to infer the parameter given data]

 [likelihood-free inference and information about ABC SMC]

 [model comparison, ABC SMC for model comparison]

 Bayesian inference approaches put Bayes Rule in the centre:

 \begin{align}
    \label{eq:bayes}
    p(\theta|D) = \frac{p(D|\theta)p(\theta)}{P(D)}
\end{align}

@@

where $l$ represents the likelihood function. In words, Eqn \ref{eq:bayes2} can be described as `posterior is proportional to the product of likelihood and prior', and here posterior of parameters is our target. Prior $p(\theta)$ should express our initial believes or estimates on the parameter values before observing the data and conducting experiments, and likelihood function $l(D|\theta)$ is used to describe how well the parameter set $\theta$ explains the data $D$, in probability numbers. Once the data is given, $\theta$ is the only parameter in the likelihood function. Some optimisation-based methods try to find an estimation of parameter set $\theta$ s.t. maximise the likelihood, such methods will struggle when facing a problem where it is hard to specify an analytical expression of likelihood function, let alone the derivative.

Based on Eqn \ref{eq:bayes2}, to find the posterior of parameter set, prior can be set according to the problem context and our understanding and cover a reasonable ranges; computational methods can be used to approximate likelihood function, e.g. Markov Chain Monte Carlo (MCMC) \cite{ref:MCMC}. As another alterative solution, Approximate Bayesian computation (ABC) approaches do not focus on the evaluation of likelihood, but focused on simulating the data from the given model and observed data, then used the simulated data to approximate the desired posterior. ABC-rejection \cite{ABC_rejection} is a simple example of the ABC process, and further algorithms e.g. sequential Monte-Carlo (ABC SMC)\cite{Toni} are proposed to improve the efficiency.

@@

In our study of the regeneration model, full trajectories of the dependent variables are used as the data to be compared, as our goal is to find a model and a parameter set that can intuitively recover the real process as much as possible. The mean values at each time point for each dependent variable is calculated to form $D$, and for each sample, the ODE is accumulatively solved at the same time points and generate the simulated trajectories $D^*$. Using other statistics to summary the raw data are also possible bur care must be taken as the choice of summary statistics can largely affect the resultant posterior \cite{summaryD, summaryD2}.

Instead of fixing the threshold value, the algorithm can be more adaptive by using a sequence of deceasing threshold values $\epsilon_t$, and derive a new population (called generation) of accepted sampled from the previous accepted samples (called particles). An example is ABC SMC \cite{Toni}, which can significantly reduce the required samples compared to ABC-rejection \cite{ref:disease}:

@@

An illustrative graph of the ABC SMC procedures can be found in Figure \ref{fig:smc}. In implementations, many factors can affect the efficiency and goodness of fit, some of them are discussed in the Chapter 4. Rather than a pre-fixed threshold schedule, median or quantile based epsilon schedule can be save total sampling size and thus improve acceptance rates of particles \cite{threshold}. Some other improvements include using adaptive population size for each generation \cite{population}, using adaptive distance function \cite{ref:adpt_dis} and switching kernels \cite{ref:kernel}, etc. In our implementations these options were tried when applicable and suitable to compare the behaviours of the algorithm.

@@

In addition, ABC SMC is also capable of model selection tasks \cite{model_compare}. Instead of draw samples from prior of a single model, we can draw samples from multiple models $m_i$ and approximate $p(m_i, \theta|D)$, which is the joint probability distribution of model $m_i$ and its accepted particles $\theta$. In each generation the marginal distribution $p(m_i|D)$ can be calculated as the accepted ${\theta}$ have been obtained, thus models can be ranked according to their probabilities.

@@

 [existing tools and software for this task]

 [workflow and developing process]

 As we decided to apply ABC SMC on our regeneration models for its high-dimensional parameter estimation and possibly comparison of different models, a code development and experiments workflow was expected to be built. The ABC SMC algorithm needs a whole set of mathematical environments (e.g. class of distributions, models and particles) and closely linked sampling and fitting algorithms (e.g. KDE fitting and Gaussian sampling), thus suitable platforms with related packages are preferred. Regarding this, several packages in \verb|R| and \verb|Python| were studied.

 A software packages that integrally implement ABC SMC is preferred, and more options in the settings of the algorithm, more customisable features can make it better. Moreover, as the ABC SMC itself is a computationally intensive task that consists of massive parallel potentials, an ideal software is expected to support as least one kind of parallel options e.g. multi-core, GPU accelerating and clusters. Otherwise, a modification or rewrite of the software package are expected to produce a reliable input-output framework for our parameter estimation and model selection experiments.

 Preliminarily \verb|pyABC| \cite{ref:pyabc} in \verb|Python| was chosen for implementations. \verb|pyABC| supports multiple types of parallelisation, e.g., multicore processing and distributed cluster, and supports resume stored runs, which made our experiments easier. It is well documented and flexible in customisations of the algorithm.

@@

 Additionally, some other softwares were also need for solving ODE, results analysis and visualisations. Some software and hardware environments are listed in Appendix A.

>>>>

%  [describe the process and how cells and cytokines are involved]

Compared to mammals, zebrafish can regenerate spinal cord after injury, and possibly recover swimming function. They are widely used in regenerative ability researches due to this ability. Experiments found that although being cut off, the axonal of nerve cells can regenerate and bridge across the lesion site, which critically determined whether zebrafish can regain the swimming function \cite{axonal}. Tsarouchas et al. \cite{ref:Tsarouchas} studied the the complex interactions happen at the lesion site. The inflammation affects the regeneration, as if it is promoted, it accelerates the axonal regeneration; if it is inhibited, it reduces the regeneration. Further experiments showed that two kinds of cytokines, il-1$\beta$ and tnf-$\alpha$, are related the inflammation, and they are dynamically controlled by immunes cells, i.e. cytokines works as media to control the axonal regeneration. Cytokines are small proteins, work as messenger to pass signal across different cells, which are similar to hormones and neurotransmitter.

An interaction map can be drawn, according to the evidences found by Tsarouchas et al. \cite{ref:Tsarouchas}. After injury, a massive immune response was observed, with the increasing numbers of several kinds of cells: macrophage, neutrophil. This dynamical system consists of complex interactions between cells and cytokines, For example, il-1$\beta$ promotes macrophages' increase, and as a major source, more macrophages leads to more expression of tnf-$\alpha$ mRNA. Consequently as media, tnf-$\alpha$ and il-1$\beta$ plays determinative roles in the axonal regeneration, where generally il-1$\beta$ promotes the regeneration at early stages, but later has an inhibition effect; tnf-$\alpha$ promotes the regeneration all the time. The presence of the cytokines also affects the number of immune cells, while as sources, more cells lead to more expression of cytokines. Interactions between different kinds of cells are also considered in the system.

This dynamic system is suitable for mathematical modelling, where a set of equations can be written to represent the dynamics of different objects e.g. cells and cytokines. Our interest lies in the modelling of the regeneration period of the dynamic system, and we wish to find that if we can quantitatively study the interactions iniside the system, based on the proposed hypothesis \cite{ref:Tsarouchas}, and furthermore based our own hypothesis which may include different interactions.

@@

%  [how this can be modelled as dynamic systems]

%  [general topics of mathematical models: how to build model according to interactions; how to calibrate/ parameterise the model; how to solve the model; dynamics; how to evaluate the model]

Mathematical models can express real-life problems in mathematical forms, and provide qualitative or quantitative inside-look of the question through various mathematical tools and techniques. Mathematical models can be used for prediction, classification, testing our understanding or hypothesis and many other tasks, thus they are widely used to solve practical problems. However, for a model that describes the zebrafish spinal cord regeneration process, we were less interested to the model's predictive functionality; the main focus is on the explore of uncertainty in parameter estimates and comparison with alternative models. In this case, we wish to use modelling techniques to implement inference methods, test our hypothesis and hopefully find and compare well-fitted models.

According to our understanding of the biological mechanism in the regeneration, hypothesis should firstly proposed in a way that can represent the real  mechanism as much as possible. Usually a real-life problem, e.g. the spinal cord regeneration, consists of complex dynamics and interactions with many other environmental factors to be considered; an elaborate description of the real cases using a overly complex mathematical model sometimes can be impossible; assumptions and hypothesis are often proposed to make necessary simplification and abstract for a final reasonably detailed model. Some qualitative models are simple but illustrative for the real-life observed patterns, e.g. Lotka-Volterra (L-V) equation for predator and prey models; they are usually analytically solvable and some properties are used to make predictions (e.g. steady states in ecosystem). If a more precise model is desired, usually more effects or interactions are considered and added as terms of equations to a re-constructed model, e.g. considering environmental factors, a changing environmental capacity and predation from other species in L-V model.

Models like L-V equation are mostly analytically solvable when analysing properties of the model, however, analytical solution for many other general models are not feasible, because of more complex dynamics, or the model itself is not a deterministic model. In the areas where solutions are not analytically feasible, computational techniques are usually applied for approximate solutions. Solvers for different models had been proposed to simulate the system or find roots for some spacial points, such as Explicit Runge-Kutta \cite{dormand1980family} and LSODA.

For the regeneration process that we wanted to model, differential equations are our preference. Non-linear ordinary differential equations are used in this dissertation, where the only independent variable is time $t$, denoting the post-lesion time (hours). Further models can also consider other models in other forms, e.g. stochastic models and partial differential equations. For ordinary differential equations (ODE) system, there are several available computational solvers using different algorithms, implemented in MATLAB\footnote[1] {\url{https://uk.mathworks.com/help/matlab/math/choose-an-ode-solver.html}}, Python\footnote[2]{\url{https://docs.scipy.org/doc/scipy/reference/generated/scipy.integrate.solve_ivp.html}} and many other languages. These solvers help to simulate the trajectories of the dependent variables, given initial values and a set of fixed parameters. The goodness of `fit' then can be measured by compared the simulated data and observed real data.

Often further researches on the modelled problems require the model to be deterministic on its parameters. Parameters in a dynamic system models usually have their practical significance, e.g. describing reaction speed, environmental capacity or the birth rate. There are many approaches to derive these parameters from data or experience. For a popular topic and a widely used models, the parameters' values can be read from previous literatures; however in many others cases, we need to find parameter's values specifically for our model, and numerical values for parameters are mostly preferred, as they suitable for practical analysis and comparison.

To find the best parameters' values given observed data, either optimisation-based approaches (e.g. least-square (LS) fit) or inference approaches can be used. Optimisation-based approaches may face two major drawbacks: over-fitting and local optima \cite{ref:abcsysbio}. Bayesian inference approaches gain popularity in recent years, which can relief the concerns in over-fitting and provide a measurement for the uncertainty in the inferred values. Local optima can still be a concern, and many efforts have be proposed for a more through explore of the parameter space to avoid this.

@@

%  [how to parameterise the model]

%  [how to infer the parameter given data]

%  [likelihood-free inference and information about ABC SMC]

%  [model comparison, ABC SMC for model comparison]

 Bayesian inference approaches for parameter estimation put Bayes Rule in the centre:

 \begin{align}
    \label{eq:bayes}
    p(\theta|D) = \frac{p(D|\theta)p(\theta)}{p(D)}
\end{align}

@@

where $l$ represents the likelihood function. In other words, Eqn \ref{eq:bayes2} can be described as `posterior is proportional to the product of likelihood and prior', and here posterior of parameters is our target in parameter estimation task. Prior $p(\theta)$ should express our initial believes or estimates on the parameter values before observing the data and conducting experiments; likelihood function $l(D|\theta)$ is used to describe how well the parameter set $\theta$ explains the data $D$, in probability values. Once the data is given, $\theta$ is the only parameter in the likelihood function. Some optimisation-based methods try to find estimates of parameter set $\theta$ s.t. maximise the likelihood, however, such methods will struggle when facing a problem where it is hard to specify an analytical expression of likelihood function, let alone the derivative of it.

Based on Eqn \ref{eq:bayes2}, to find the posterior of parameter set, prior can be set according to the problem context and our understanding and cover a reasonable ranges; computational methods can be used to approximate likelihood function, e.g. Markov Chain Monte Carlo (MCMC) \cite{ref:MCMC}. As an alterative solution, Approximate Bayesian computation (ABC) approaches do not focus on the evaluation of likelihood function, but focused on simulating the data from the given model and observed data, then used the simulated data to approximate the desired posterior. ABC-rejection \cite{ABC_rejection} is a simple example of the ABC process, and further algorithms e.g. sequential Monte-Carlo (ABC SMC)\cite{Toni} were proposed to improve the efficiency.

@@

In our study of the regeneration model, full trajectories of the dependent variables are used as the data to be compared, as our goal is to find a model with a parameter set that can intuitively recover the real process as much as possible. The mean values at each time point for each dependent variable is calculated to form $D$, and for each sample, the ODE is accumulatively solved at the same time points and generate the simulated trajectories $D^*$. Using other statistics to summary the raw data are also possible but care must be taken as the choice of summary statistics can largely affect the resultant posterior \cite{summaryD, summaryD2}.

Instead of fixing the threshold value, the algorithm can be more adaptive by using a sequence of deceasing threshold values $\epsilon_t$, and derive new populations (called generation) of samples from the accepted samples (called particles) in previous population. An example is ABC SMC \cite{Toni}, which can significantly reduce the required samples compared to ABC-rejection \cite{ref:disease}:

@@

An illustrative graph of the ABC SMC procedures can be found in Figure \ref{fig:smc}. In implementations, many factors can affect the efficiency and goodness of fit, some of them are discussed in the Chapter 4. Rather than a pre-fixed threshold schedule, median or quantile based epsilon schedule can save the total sampling size and thus improve acceptance rates of particles \cite{threshold}. Some other improvements include using adaptive population size for each generation \cite{population}, using adaptive distance function \cite{ref:adpt_dis} and switching kernels \cite{ref:kernel}, etc. In our implementations these options were tried and compared when applicable and suitable.

@@

In addition, ABC SMC is also capable of model selection tasks \cite{model_compare}. Instead of draw samples from prior of a single model, we can draw samples from multiple models $m_i$ and approximate $p(m_i, \theta|D)$, which is the joint probability distribution of model $m_i$ and its accepted particles $\theta$. In each generation the marginal distribution $p(m_i|D)$ can be calculated as the accepted ${\theta}$ is known, thus models can be ranked according to their probabilities.

@@

%  [existing tools and software for this task]

%  [workflow and developing process]

 As we decided to apply ABC SMC on our regeneration models for its high-dimensional parameter estimation and comparison of different models, a code development and experiments workflow was expected to be built. The ABC SMC algorithm needs a whole set of mathematical environments (e.g. class of distributions, models and particles) and closely linked sampling and fitting algorithms (e.g. KDE fitting and Gaussian sampling), thus suitable platforms with related software packages are preferred. Regarding this, several packages in \verb|R| and \verb|Python| were studied.

 A software packages that integrally implement ABC SMC is preferred, and more options in the settings of the algorithm, more customisable features make it better. Moreover, as the ABC SMC itself is a computationally intensive task that contains massive parallel potentials, an ideal software package is expected to support as least one kind of parallel options e.g. multi-core, GPU accelerating and clusters. Otherwise, a modification or rewrite of the software package are expected to produce a reliable input-output framework for our parameter estimation and model selection experiments, where much more work would be introduced.

 Preliminarily \verb|pyABC| packages \cite{ref:pyabc} in \verb|Python| was chosen for implementations. \verb|pyABC| supports multiple types of parallelisation, e.g., multicore processing and distributed cluster, and supports resume stored runs, which made our experiments easier. It is well documented and flexible in customisations of the algorithm.

@@

 Additionally, some other softwares were also need for solving ODE, results analysis and visualisations. These software and hardware environments are listed in Appendix A.

### 2020-08-10 introduction/introduction.tex
C1 and C2 read

 [background and motivation]

 [what is done]

While we are interested in the tissue regeneration among different species, mathematical models and techniques can help us to resolve complex interactions of cells and chemicals in a biological system. The objective process in this dissertation is an example of complex biology system found in tissue regeneration. Zebrafish has the ability to regenerate its spinal cord after injury; at the injury-site various immune cells are presented and complex interactions are found while they are regulating inflammation and repair.

Our interest lies in the modeling of the regeneration process that happens in the injury-site, especially in the parameter estimation and comparison of different models. According to available researches \cite{ref:Tsarouchas}, a mathematical model is proposed but the addressed interactions remains to be precisely defined, and quantitatively examined in experiments. However, we would like to explore the the `best fit' if the model which comes with a set of parameter estimates given the observed data. Also, we wish to compare different it with different models and possibly find the best model to describe the data without losing biological and mathematical properties in the regeneration process.

Fo the parameter estimation tasks, Bayesian inference techniques can help in the problems of over-fitting and local optima \cite{ref:abcsysbio}, compared to optimisation based methods. Moreover, approximate inference methods (Approximate Bayesian Computation, ABC) make likelihood-free inference possible, which is suitable for our task as writing down likelihood as a function expression can be hard.

Such methods introduce computation-intensive sampling-rejection pattern, where massive number of samples are drawn, followed by computations. As the sampling process is independent from each other in a certain period, the parallel execution becomes feasible and is expected to accelerated the parameter inference in a large scale.

This dissertation includes the implementations and experiments of parameter estimation using ABC SMC (sequential Monte-Carlo) method, and discussions on the results. Model comparison results are also included using the same method. Additionally, a performance test illustrates the strong-scaling performance of the algorithm.

>>>>

%  [background and motivation]

%  [what is done]

While we are interested in the tissue regeneration among different species, mathematical models and techniques can help us to resolve complex interactions and effects of cells and chemicals in a biological system. The objective in this dissertation is an example of complex biology system found in tissue regeneration: zebrafish spinal cord regeneration. Zebrafish has the ability to regenerate its spinal cord after injury; at the lesion site various immune cells are presented and complex interactions are found while they are regulating inflammation and repair.

Our interest lies in the modeling of the regeneration process that happens around the lesion site, especially in the model parameter and comparison of different models. According to available researches \cite{ref:Tsarouchas}, a mathematical model was firstly proposed but the addressed interactions remains to be precisely defined, and quantitatively examined in experiments. However, we would like to explore the the `best fit' of the model which comes with a set of parameter estimates, given the observed data. Also, we wish to compare different it with other alternative models and possibly find the best model to describe the observed data. In these steps, models are expected to retain mathematical and biological significance, where differential equations are suitable to describe this system.

Fo the parameter estimation tasks, Bayesian inference techniques can help in the problems of over-fitting and local optima \cite{ref:abcsysbio}, compared to optimisation-based methods. Moreover, approximate inference methods (Approximate Bayesian Computation, ABC) make likelihood-free inference possible, which is suitable for our task as writing down likelihood as a function expression for the dynamic system can be hard.

Such methods introduce computation-intensive sampling-and-test patterns, where massive number of samples are drawn and followed by complex computations. As the sampling is independent from each other in a certain period, the parallel implementation of the algorithm becomes feasible and is expected to accelerate the parameter inference in a large scale when using HPC resources.

This dissertation includes the implementations and experiments of parameter estimation using ABC SMC (sequential Monte-Carlo) method, and discussions on the results. ABC SMC was also applied in model comparison. Additionally, performance experiments illustrated the strong-scaling performance of the algorithm.
